A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-05-28 02-ceu.tex
(*) final version

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that toggles a LED on radio receive,
           terminating on a button press always with the LED~off.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

@@

\subsection{External and Internal Events}
\label{sec.ceu.evts}

@@

\subsubsection*{External events and discrete time}

@@

\begingroup
\begin{figure}[h]
\centering
\begin{tikzpicture}[x=1em,y=1em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\footnotesize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Real World)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(\CEU)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
\vskip-.6em
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

@@

\subsubsection*{Internal events as subroutines}

@@

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A ``subroutine'' that increments its argument.},
  label={lst.sub},
]
event int* inc;       |\C{declares subroutine "inc"}|
par/or do
    var int* p;
    every p in inc do |\C{implements "inc" through event iterator}|
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);     |\C{calls "inc"}|
    emit inc(&v);     |\C{calls "inc"}|
    _assert(v==3);    |\C{asserts result after the two returns}|
end
\end{lstlisting}
%\vskip-\baselineskip

@@

@@

\begingroup
\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\small{\ax\ Accesses to \code{x} are never concurrent.}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
input void A;
|\C{(empty line)}|
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

\begin{figure}[t]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
par/or do
   var _msg_t msg;
   $\T<\cdots>$   |\C{prepare msg}|
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   $\T<\cdots>$
end
|\C{(empty line)}|
\end{lstlisting}
\small{\ax\ Local resource finalization.}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
par/or do
   var _FILE* f;
   finalize
      f = _fopen($\dots$);
   with
      _fclose(f);
   end
   _fwrite($\dots$, f);
   await A;
   _fwrite($\dots$, f);
with
   $\T<\cdots>$
end
\end{lstlisting}
\hskip1em
\begin{minipage}[t]{.94\columnwidth}
\small{\bx\ External resource finalization.}
\end{minipage}%
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
\label{lst.fin.ceu}
\end{figure}

>>>>

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that toggles a LED on receive,
           terminating on a button always with the LED~off.},
  label={lst.ceu},
]
input void BUTTON;
input void RADIO_RECV;
par/or do
    await BUTTON;
with
    finalize with
        _led(0);
    end
    loop do
        _led(1);
        await RADIO_RECV;
        _led(0);
        await RADIO_RECV;
    end
end
\end{lstlisting}

\vspace{-0.5cm}
\subsection{External and Internal Events}
\label{sec.ceu.evts}

@@

\subsubsection{External Events and Discrete Time}

@@

\begingroup
\begin{figure}[b!]
\centering
\begin{tikzpicture}[x=1em,y=1em,
  font=\footnotesize\sffamily,>=stealth,
  time/.style={font=\scriptsize\sffamily},
  cross/.style={inner sep=0pt,outer sep=0pt},
  evt/.style={fill=white,inner sep=1.85pt,outer sep=0pt},
  rng/.style={fill=white,inner sep=1pt,outer sep=0pt,
    font=\ttfamily\footnotesize}]
  \draw[->]
  (0,0)
    node[align=right,anchor=east,xshift=-.5em]
        {Continuous\\Timeline\\(Real World)}
    node[anchor=north west,xshift=-.5em,time]
        {0ms000us}
    node[anchor=north west,xshift=8.75em,time]
        {17ms521us}
  -- (18.5,0)
    node[anchor=north east,xshift=.1em,time]
        {30ms000us};
  \node[cross] at (0,0){x};
  %%
  \coordinate(Ext) at ($(0,0)!.5!(0,-5)$);
  \coordinate(A) at (Ext-|4,0);
  \coordinate(B) at (Ext-|5.5,0);
  \coordinate(C) at (Ext-|9,0);
  \coordinate(D) at (Ext-|15,0);
  %%
  \draw[semithick,dotted] (0,1-|A) -- (0,-6-|A);
  \draw[semithick,dotted] (0,1-|B) -- (0,-6-|B);
  \draw[semithick,dotted] (0,1-|C) -- (0,-6-|C);
  \draw[semithick,dotted] (0,1-|D) -- (0,-5-|D);
  \node[evt](Aevt) at (A) {\texttt{\textbf{A}}};
  \node[evt](Bevt) at (B) {\texttt{\textbf{B}}};
  \node[evt](Cevt) at (C) {\texttt{\textbf{C}}};
  \node[evt](Devt) at (D) {\texttt{\textbf{D}}};
  \node[anchor=east,xshift=-.5em] at (A){\emph{External Input}};
  %%
  \draw[-]
  (0,-5)
    node[align=right,anchor=east,xshift=-.5em]
        {Discrete\\Timeline\\(\CEU)}
  -- (18.5,-5);
  \node[cross](P0) at (0,-5){x};
  \node[cross](P1) at (4.75,-5){x};
  \node[cross](P2) at (10,-5){x};
  \node[cross](P3) at (12.5,-5){x};
  \node[cross](P4) at (14.5,-5){x};
  \node[cross](P5) at (15.5,-5){x};
  \node[cross](P6) at (18.5,-5){x};
  \node[rng] at ($(P0)!.5!(P1)$){boot-0};
  \node[rng] at ($(P1)!.5!(P2)$){A-1};
  \node[rng] at ($(P2)!.5!(P3)$){B-2};
  \node[rng] at ($(P3)!.5!(P4)$){C-3};
  \node[rng] at ($(P5)!.5!(P6)$){D-4};
  \draw[decorate,decoration={brace,raise=-1pt,mirror,amplitude=2pt}]
  ($(P4)+(0,-.5)$) -- node[below,time] {Idle period} ($(P5)+(0,-.5)$);
  %%
  \begin{scope}[shorten >=1.5pt]
  \draw[->] (Aevt) -- (P1);
  \draw[->] (Bevt) -- (P2);
  \draw[->] (Cevt) -- (P3);
  \draw[->] (Devt) -- (P5);
  \end{scope}
\end{tikzpicture}
\vskip-.6em
%\includegraphics[width=\columnwidth]{tick_min}
\caption{The discrete notion of time in \CEU.}
\label{fig.ticks}
\end{figure}
\endgroup

@@

\subsubsection{Internal Events as Subroutines}

@@

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A ``subroutine'' that increments its argument.},
  label={lst.sub},
]
event int* inc;       |\C{declares subroutine "inc"}|
par/or do
    var int* p;
    every p in inc do |\C{implements "inc" as event iterator}|
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);     |\C{calls "inc"}|
    emit inc(&v);     |\C{calls "inc"}|
    _assert(v==3);    |\C{asserts result after the two returns}|
end
\end{lstlisting}
%\vskip-\baselineskip

@@

\begingroup
\begin{figure}[b!]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
input void A;
input void B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\small{\ax\ Accesses to \code{x} are never concurrent.}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
%\begin{lstlisting}
\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
input void A;
|\C{(empty line)}|
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end

@@

%\vspace{2cm}
\begin{figure}[b!]
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=right,basicstyle=\ttfamily\small]
par/or do
   var _msg_t msg;
   $\T<\cdots>$   |\C{prepare msg}|
   finalize
      _send(&msg);
   with
      _cancel(&msg);
   end
   await SEND_ACK;
with
   $\T<\cdots>$
end
|\C{(empty line)}|
\end{lstlisting}
\small{\ax\ Local resource finalization.}
\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[xleftmargin=1.75em,basicstyle=\ttfamily\small]
par/or do
   var _FILE* f;
   finalize
      f = _fopen($\dots$);
   with
      _fclose(f);
   end
   _fwrite($\dots$, f);
   await A;
   _fwrite($\dots$, f);
with
   $\T<\cdots>$
end
\end{lstlisting}
\hskip1em
\begin{minipage}[t]{.94\columnwidth}
\small{\bx\ External resource finalization.}
\end{minipage}%
\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{%
\CEU enforces the use of finalization to prevent dangling pointers and memory leaks.}
\label{lst.fin.ceu}
\end{figure}
